The proof equivalence problem for multiplicative linear logic is \textsc{pspace}-complete v0.3

### 2013-09-13 main/s_level_0.tex
Fixes, additions, and pictures for Level-0

\begin{proof}
%
It will be shown by induction on the number of $\bot$-formulae in $\Gamma$ that every proof net for $\Gamma$ belongs to one of two equivalence classes.
%
For the base case, the smallest inhabited sequent with two tensor-formulae is the following.
\[
	\1,\1,\1,\bot\tn\bot,\bot\tn,\bot
\]
It has two equivalence classes of 12 proof nets each.

For the inductive step, let $\Gamma$ be the following sequent.
\[
	\Delta,A\tn\named a\bot,\named x\1
\]
There are two cases: 1) where $A$ is a tensor-formula, and 2) where $A$ is $\bot$ and where, for the induction hypothesis to apply, $\Delta$ contains at least two tensor-formulae.
%
For both cases, it will be shown that any net $\links$ for $\Gamma$ is equivalent to a net $\links'$ where $\named a\bot$ connects to $\named x1$, and is the only link to do so.
%
This reduces equivalence on $\Gamma$ to equivalence on $\Delta,A$ in case 1, and on $\Delta$ in case 2, so that the induction hypothesis applies.

Let $\named a\bot$ connect to $\named y\1$ in $\Delta$, and let $\named c\bot$ be a subformula of $A$, which means that for case 2, $A=\named c\bot$.
%
Then $\links'$ is obtained by adjusting $\links$ as follows.
\begin{itemize}

	\item
Let $\edge cz$, i.e.\ $\named z\1$ is the target of the jump from $\named c\bot$.	
%
Ensure that $\named z\1$ is the only target shared between jumps in $A$ and in $\Delta$, by moving any other such jump from $\Delta$ to $\named z\1$.

	\item
If there are multiple links connecting to $\named x\1$, select one $b-x$ for some $\named b\bot$ in a tensor-formula $B$.
%
Re-attach the others to the target of another jump out of $B$, of which there must be at least one.

	\item
Since $A$ is only connected via $\named z\1$, there is a link $\edge dz$ connecting $B$ to $A$ (though $\named d\bot$ is not necessarily a subformula of $B$).
%
Re-attach $\named d\bot$ to $\named y\1$, then $\named a\bot$ to $\named x\1$, and $\named b\bot$ to $\named z\1$.

\end{itemize}

\end{proof}

\begin{proposition}
\label{prop:level0 may-connect path}
%
In a proof net for a 1-alternation sequent a link $\edge{\named a\bot}{\named b\1}$ can be permuted to $\edge{\named a\bot}{\named c\1}$ if and only if there is a path in the net from $b$ to $c$ not passing through $a$.
%
\end{proposition}

@@

	\item
The first $\1$ in $\Gamma$ is named $N$, the remaining ones are named with the numbers $n+1,\dotsc,m$.

	\item
A $\bot$-formula in $A_i$ is named by a pair $(i,k)$, where $k=N$ for the first $\bot$-formula in each $A_i$, and for the remaining $\bot$-formulae in all $A_i$, each $k$ is a distinct number in $n+1,\dotsc,m$.

\end{itemize}
%
The naming scheme suggests a linking for $\Gamma$, where $\edge{(i,k)}{k}$ for each $\bot$-formula; i.e\ the first $\bot$-subformula of each tensor-formula connects to $\named N\1$, while other $\bot$-subformulae connect uniquely to the remaining $\1$-subformulae.

A net for $\Gamma$ is interpreted as a combinatorial permutation (an automorphism on $\{1,\dotsc,m\}$) as follows.
%
\begin{definition}
To a proof net $\links$ for a 1-alternation sequent $\Gamma$ named as above, associate the \emph{permutation} $\prm:\{1,\dotsc,m\}\to\{1,\dotsc,m\}$ given by:
\[
	\prm(k) = 
	\begin{cases}
		i				& \text{ if $(i,k)$ may connect to $N$; otherwise,}
	\\	j				& \text{ where $\edge{(i,k)}j$.}
	\end{cases}
\]
The \emph{parity} of $\links$ is the parity of its permutation.
\end{definition}

To see that $\prm$ is injective, consider the following.
\begin{itemize}
	\item The domains of $i$ and $j$ are $1,\dotsc,n$ and $n+1,\dotsc,m$ respectively, and hence disjoint.
	\item Exactly one $\bot$-formula in each $A_i$ may connect to $N$ because of connectedness and acyclicity, since if a $\bot$-formula may connect to $N$ it has a path to $N$ (Proposition~\ref{prop:level0 may-connect path}).
	\item If two $\bot$-formulae have the same target, which means they are in different tensor-formulae, at least one may connect to $N$ via the other tensor-formula, which must have a path to $N$ by the above.
\end{itemize}

\begin{proposition}
\label{lem:level0 min binary}
Re-attaching a jump in a net $\links$ preserves its parity. 
\end{proposition}

\begin{proof}
Let $\links$ be a net for $\Gamma$, with $\Gamma$ named as above, and let the link $\edge{(i,k)}x$ in $\links$ re-attach as $\edge{(i,k)}y$, forming $\links'$.
%
There are two cases, depending on whether $(i,k)$ may connect to $N$.
%
If so, using Proposition~\ref{prop:level0 may-connect path}, the re-wiring preserves which $\bot$-formulae may connect to $N$, since for any path to $N$ via $\edge{(i,k)}x$ in $\links$ there is a path to $N$ via $\edge{(i,k)}y$.
%
Then the permutation of $\links'$ is that of $\links$.

If $(i,k)$ may not connect to $N$, let the path from $x$ to $y$ run via the following $\bot$- and $\1$-vertices.
\[
	x=x_1, (i_1,j_1), (i_1,k_1), x_2, (i_2,j_2), \dotsc, (i_n,k_n), x_{n+1}=y 	
\]
Note that the $\bot$-formulae $(i_a,j_a)$ may connect to $N$.
%
On the relevant domain, this gives the following permutation for $\links$.
\[
\left(\begin{array}{ccccccc}
	j_1 & \dotso & j_n &  k  & k_1 & \dotso & k_n \\
	i_1 & \dotso & i_n & x_1 & x_2 & \dotso & x_{n+1}
\end{array}\right)
\]
In $\links$, where $(i,k)$ connects to $y$, the $\bot$-formulae $(i_a,k_a)$ may connect to $N$, giving the following permutation.
\[
\left(\begin{array}{ccccccc}
	j_1 & \dotso & j_n &    k    & k_1 & \dotso & k_n \\
	x_1 & \dotso & x_n & x_{n+1} & i_1 & \dotso & i_n
\end{array}\right)
\]
The parity of both permutations is the same if and only if the relative permutation, below, is even.
\[
\left(\begin{array}{ccccccc}
	i_1 & \dotso & i_n & x_1     & x_2 & \dotso & x_{n+1} \\
	x_1 & \dotso & x_n & x_{n+1} & i_1 & \dotso & i_n
\end{array}\right)
\]
This is the case, as it is obtained by the exchange of $x_a$ and $i_a$ for each $a\leq n$, and subsequently the exchange of $x_{n+1}$ and each $i_a$ in turn.

\end{proof}

>>>>

%
%\begin{figure}[p]
%\[
%\begin{array}{ccc}
%	\netA 1213 & \perm	& \netA 3213 \\ \\[-6pt] \vperm && \vperm \\ \\[-6pt]
%	\netA 1223 &		& \netA 3212 \\ \\[-6pt] \vperm && \vperm \\ \\[-6pt]
%	\netA 1323 &		& \netA 3112 \\ \\[-6pt] \vperm && \vperm \\ \\[-6pt]
%	\netA 1321 &		& \netA 3132 \\ \\[-6pt] \vperm && \vperm \\ \\[-6pt]
%	\netA 2321 &		& \netA 2132 \\ \\[-6pt] \vperm && \vperm \\ \\[-6pt]
%	\netA 2331 & \perm	& \netA 2131
%\end{array}
%\qquad
%\begin{array}{ccc}
%	\netA 1312 & \perm	& \netA 1332 \\ \\[-6pt] \vperm && \vperm \\ \\[-6pt]
%	\netA 2312 &		& \netA 1232 \\ \\[-6pt] \vperm && \vperm \\ \\[-6pt]
%	\netA 2313 &		& \netA 1231 \\ \\[-6pt] \vperm && \vperm \\ \\[-6pt]
%	\netA 2113 &		& \netA 3231 \\ \\[-6pt] \vperm && \vperm \\ \\[-6pt]
%	\netA 2123 &		& \netA 3221 \\ \\[-6pt] \vperm && \vperm \\ \\[-6pt]
%	\netA 3123 & \perm	& \netA 3121
%\end{array}
%\]
%\caption{The two equivalence classes of nets for $\1,\1,\1,\bot\tn\bot,\bot\tn,\bot$}
%\label{fig:2x12 nets}
%\end{figure}

\begin{proof}
%
It will be shown by induction on the number of $\bot$-formulae in $\Gamma$ that every proof net for $\Gamma$ belongs to one of two equivalence classes.
%
For the base case, the smallest inhabited sequent with two tensor-formulae is the following.
\[
	\1,\1,\1,\bot\tn\bot,\bot\tn,\bot
\]
It has two equivalence classes, of 12 proof nets each.
%
Apart from listing these exhaustively, this can also be shown by using the proof of the inductive step, below, to reduce the base case to that of the sequent $\1,\1,\bot\tn\bot$, which has two singleton equivalence classes.

%, displayed in Figure~\ref{fig:2x12 nets}.

For the inductive step, let $\Gamma$ be the following sequent.
\[
	\Delta,A\tn\named a\bot,\named z\1
\]
There are two cases: 1) where $A$ is a tensor-formula, and 2) where $A$ is $\bot$ and where, for the induction hypothesis to apply, $\Delta$ contains at least two tensor-formulae.
%
For both cases, it will be shown that any net $\links$ for $\Gamma$ is equivalent to a net $\links'$ where $\named a\bot$ connects to $\named z1$, and is the only link to do so.
%
This reduces equivalence on $\Gamma$ to equivalence on $\Delta,A$ in case 1, and on $\Delta$ in case 2, so that the induction hypothesis applies.

In constructing $\links'$, since the proof net must be connected, there is a path from $a$ to $z$.
%
We will consider only the jumps on this path, not any other edges.
%
There are four cases.

\begin{enumerate}[label=\roman*.]
	\item
The path consists of exactly the jump $\edge az$.
%
Let $\edge by$ be a jump from a $\named b\bot$ in $A$; then $\links'$ is obtained from $\links$ by changing: $\links'(e)=y$  for every $\named e\bot$ such that $\links(e)=z$ and $e\neq a$.

	\item
The path starts with the jump $\edge ax$ (and $x\neq z$).
%
Let $\edge by$ be a jump from a $\named b\bot$ in $A$, and let the path end with the jumps $\edge wc$ and $\edge dz$, where $\named c\bot$ and $\named d\bot$ are in the same tensor-formula $B$.
%
Then $\links'$ is obtained from $\links$ by changing: $\links'(a)=z$, and $\links'(e)=w$ for every $\named e\bot$ such that $\links(e)=z$ and $e\neq a$, including $d$.
%
These changes are illustrated as permutations below.
%
\[
	\netB xywz \perm \netB zywz \perm \netB zywy
\]

	\item
The path consists of exactly one jump $\edge bz$ from a $\named b\bot$ in $A$ (and $b\neq a$).
%
Let $\links(a)=y$.
%
Choose a jump $\edge cw$ from $\named c\bot$ in $A$ such that there is another $\edge dw$ from $\named d\bot$ in a formula $B$ (not excluding the possibilities $c=b$ and $c=a$).
%
Let $\edge ex$ be a further jump from $\named e\bot$ in $B$.
%
Then $\links'$ is obtained from $\links$ by changing: $\links'(a)=z$, $\links'(d)=x$, $\links'(e)=y$, and $\links'(f)=x$ for each $f$ (including $b$) such that $\links(f)=z$ and $f\neq a$.
%
These changes are exhibited as a series of permutations below, from top left to bottom left (note that the jump from $d$ moves twice).
%
\[
\begin{array}{ccccc}
	\netC yzwx & \perm & \netC yzzx & \perm & \netC yxzx
	\\ &&&& \vperm \\
	\netC zxxy & \perm & \netC zxzy & \perm & \netC yxzy
\end{array}
\]

	\item
The path starts with a jump $\edge bx$ from a $\named b\bot$ in $A$ (and $b\neq a$, $x\neq z$).
%
Let the path end with the jumps $\edge wc$ and $\edge dz$, and let $\links(a)=y$.
%
Then $\links'$ is obtained from $\links$ by changing: $\links'(c)=y$, $\links'(a)=z$, and $\links'(e)=w$ for every $\named e\bot$ such that $e\neq a$ and $\links(e)=z$, including $d$.
%
This is illustrated below.
\end{enumerate}
%
\[
	\kern-1pt \netB yxwz \perm \netB yxyz \perm \netB zxyz \perm \netB zxyw \kern-1pt 
\]
%
\end{proof}

\begin{proposition}
\label{prop:level0 may-connect path}
%
For a proof net for a 1-alternation sequent containing a link $\edge{\named a\bot}{\named b\1}$ and a formula $\named c\1$, the following are equivalent.
%
\begin{itemize}
	\item
The edge $\edge ab$ can be permuted to $\edge ac$.
	\item
There is a path from $b$ to $c$ not passing through $a$.
	\item
The path from $a$ to $c$ starts with the jump $a-b$.
\end{itemize} 
\end{proposition}

@@

	\item
One $\1$ in $\Gamma$ is named $*$, and the remaining ones with the numbers $n+1,\dotsc,m$.

	\item
A $\bot$-formula in $A_i$ is named by a pair $(i,k)$, where $k=*$ for the first $\bot$-formula in each $A_i$, and for the remaining $\bot$-formulae in all $A_i$, each $k$ is a distinct number in $n+1,\dotsc,m$.

\end{itemize}
%
The naming scheme suggests a linking for $\Gamma$, defined by $\links(i,k)=k$; i.e\ the first $\bot$ in each tensor-formula connects to $\named *\1$, while other $\bot$-subformulae connect uniquely to the remaining $\1$-subformulae.

A net for $\Gamma$ is interpreted as a combinatorial permutation (an automorphism on $\{1,\dotsc,m\}$) as follows.
%
\begin{definition}
\label{def:combinatorial permutation}
To a proof net $\links$ for a 1-alternation sequent $\Gamma$ named as above, associate the \emph{permutation} $\prm:\{1,\dotsc,m\}\to\{1,\dotsc,m\}$ given by:
\[
	\prm(k) = 
	\begin{cases}
		i				& \text{ if $(i,k)$ may connect to $*$; and}
	\\	\links(i,k)		& \text{ otherwise.}
	\end{cases}
\]
The \emph{parity} of $\links$ is the parity of its permutation.
\end{definition}

To see that $\prm$ is injective, consider the following.
\begin{itemize}
	\item The domains of $i$ and $\links(i,k)$, respectively $1,\dotsc,n$ and $n+1,\dotsc,m$, are disjoint.
	\item Exactly one $\bot$-formula in each $A_i$ may connect to $*$ because of connectedness and acyclicity, since if a $\bot$-formula may connect to $*$ it has a path to $*$ (Proposition~\ref{prop:level0 may-connect path}).
	\item If two $\bot$-formulae have the same target, which means they are in different tensor-formulae, at least one may connect to $*$ via the other tensor-formula, which must have a path to $*$ by the above.
\end{itemize}

\begin{proposition}
\label{prop:level0 min binary}
A permutation on a net $\links$ preserves its parity. 
\end{proposition}

\begin{proof}
Let $\links$ be a net for $\Gamma$, with $\Gamma$ named as above, and let the link $\edge{(i,k)}x$ in $\links$ re-attach as $\edge{(i,k)}y$, forming $\links'$.
%
There are two cases, depending on whether $(i,k)$ may connect to $*$.
%
If so, using Proposition~\ref{prop:level0 may-connect path}, the re-wiring preserves which $\bot$-formulae may connect to $*$, since for any path to $*$ via $\edge{(i,k)}x$ in $\links$ there is a path to $*$ via $\edge{(i,k)}y$.
%
Then the permutation of $\links'$ is that of $\links$.

If $(i,k)$ may not connect to $*$, let the path from $x$ to $y$ run via the following $\bot$- and $\1$-vertices.
\[
	x=x_1, (i_1,j_1), (i_1,k_1), x_2, (i_2,j_2), \dotsc, (i_n,k_n), x_{n+1}=y 	
\]
Note that the $\bot$-formulae $(i_a,j_a)$ may connect to $*$.
%
On the relevant domain, this gives the following permutation for $\links$.
\[
\left(\begin{array}{ccccccc}
	j_1 & \dotso & j_n &  k  & k_1 & \dotso & k_n \\
	i_1 & \dotso & i_n & x_1 & x_2 & \dotso & x_{n+1}
\end{array}\right)
\]
In $\links'$, since $\links'(i,k)=y$, the $\bot$-formulae that may connect to $*$ are the $(i_a,k_a)$.
%
The permutation $\prm[\links']$ is the following.
\[
\left(\begin{array}{ccccccc}
	j_1 & \dotso & j_n &    k    & k_1 & \dotso & k_n \\
	x_1 & \dotso & x_n & x_{n+1} & i_1 & \dotso & i_n
\end{array}\right)
\]
The parity of both permutations is the same if and only if the relative permutation, below, is even.
\[
\left(\begin{array}{ccccccc}
	i_1 & \dotso & i_n & x_1     & x_2 & \dotso & x_{n+1} \\
	x_1 & \dotso & x_n & x_{n+1} & i_1 & \dotso & i_n
\end{array}\right)
\]
This is the case, as it is obtained by the exchange of $x_a$ and $i_a$ for each $a\leq n$, and subsequently the exchange of $x_{n+1}$ and each $i_a$ in turn.
%
\end{proof}

\begin{theorem}
\MLL\ proof equivalence in the absence of ~$\parr$ is linear-time decidable.
\end{theorem}

\begin{proof}
For a sequent with 1 tensor-formula, the problem is syntactic equality.
%
For a sequent with 2 or more tensor-formulae, by Propositions~\ref{prop:level0 max binary} and~\ref{prop:level0 min binary} the equivalence of two nets is determined by their relative parity.
%
Following Definition~\ref{def:combinatorial permutation} the parity of a net can be read off in a single traversal of the net.
%
This yields a linear-time algorithm.
%
\end{proof}

### 2013-09-13 main/scratch.tex
Fixes, additions, and pictures for Level-0

\documentclass{robinminion}

\usepackage{amsthm}
\usepackage{enumitem}
\usepackage[round]{natbib}
\usepackage{mllequiv,willemtools}

@@

\author{Willem Heijltjes and Robin Houston}
\title{The proof equivalence problem for multiplicative linear logic is \textsc{pspace}-complete v0.3}

@@

\maketitle

\begin{abstract}
MLL proof equivalence is the problem of deciding whether two proofs are related by a series of rule permutations.
%
Previous work has shown the problem to be equivalent to a rewiring problem on proof nets, which are not canonical for full MLL due to the presence of the two units.
%
Drawing from recent work on reconfiguration problems, in this paper it is shown that MLL proof equivalence is PSPACE-complete, using a reduction from Nondeterministic Constraint Logic.
\end{abstract}

%\include{s_intro}
\include{s_background}
\include{s_level_0}
%\include{s_level_1}
%\include{s_level_2} 
%\include{s_conclusion}

\bibliographystyle{plainnat}
\bibliography{refs}

>>>>

\documentclass{article}

\usepackage{amsthm}
\usepackage[round]{natbib}
\usepackage{mllequiv,willemtools}

@@

@@
